Hardware Caches and Optimization

### 2017-06-27 tex/com.tex
Tweak stuff and fill in some blanks
Expand, revise, tweak, and run some measurements

% > In contrast to the external-memory model, algorithms in the cache-oblivious model
% > cannot explicitly manage the cache (issue block-read and block-write requests).  This
% > loss of freedom is necessary because the block and cache sizes are unknown.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf (erikcom), page 5
Cache misses take the place of \acrshortpl{io} \x{used in the \gls{emm}} because
cache-oblivious algorithms don't \alts{manage the cache, read or write cache lines}
explicitly.  This wouldn't be possible since the algorithms know neither the cache nor
the cache line size~\cite[5]{erikcom}.
% > The ideal cache uses the optimal off-line strategy of replacing the cache line whose
% > next access is furthest in the future
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1
Instead, the \gls{com} uses the optimal replacement strategy of evicting the cache line
that won't be accessed for the longest time in the future (Bélády's Algorithm).  This
\alts{
  is strangely \alts{at odds, out of touch, unrealistic} with real-world caches that don't
  know the future,
  {seems like out-of-touch, theoretical ivory-tower nonsense},
}.
% > The ideal-cache model makes the perhaps-questionable assumptions that there are only
% > two levels in the memory hierarchy, that memory is managed automatically by an optimal
% > cache-replacement strategy, and that the cache is fully associative.  We address these
% > assumptions in Section 6, showing that to a certain extent, these assumptions entail
% > no loss of generality.
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=2
%
% > LRU and FIFO replacement do just as well as optimal replacement up to a constant
% > factor of memory transfers and up a constant factor wastage of the cache.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > The same argument extends to a variety of other replacement strategies.
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46
%
% > We show that algorithms with regular complexity bounds (Equation (7.1)) (including all
% > algorithms heretofore presented) can be ported to less-ideal caches incorporating
% > least-recently-used (LRU) or first-in, first-out (FIFO) replacement policies [24, p.
% > 378].
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=51
%
% > [I]n many cases [the COM] is provably within a constant factor of a more realistic
% > cache's performance.
%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
%
% > [A]s long as the number of memory transfers depends polynomially on the cache size M,
% > then halving M will only affect the running time by a constant factor.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > Intuitively, algorithms that slow down by a constant factor when memory (M) is reduced
% > to half, are called regular.
%      -- Algorithms for Memory Hierarchies, page 196
However, \citeauthor{coa-thesis} proves that for many algorithms\footnote{%
   Those satisfying equation (7.1) in~\cite[46]{coa-thesis}.  If the number of cache
   misses incurred by the algorithm only \say{depends polynomially on the cache size
   \(M\)}, the \alts{equation, condition} is satisfied~\cite[6]{erikcom}.
   % Halving the cache size only increases the number of cache misses by a constant
   % factor?
}
it only increases cache misses by a constant factor compared to various feasible
replacement strategies%
% ~\cite[lemma 12, \pno~10]{coa-paper}.%
% ~\cite[corollary 13, \pno~10]{coa-paper}.%
~\cite[corollary 19, \pno~46]{coa-thesis}.%
% See <http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46>.
\footnote{%
   e.g. LRU, FIFO, and random replacement
}

@@

Kumar works around this by \say{[stopping] the recursion when the problem size becomes
less than a certain block size and then [using] the simple for loop implementation inside
the block}~\cite[199--201]{afmh}.  This seems to \alts{compromise, jeopardize} the
\alts{cache-obliviousness, cache-oblivious nature} of the algorithm.

Perhaps cache-oblivious algorithms should be seen as a starting point for further
optimizations that \alts{explicitly use the available information about the memory
hierarchy of the machine a program runs on, take the properties of the target machine into
account}.

@@

\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024},
               },
              },
       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
         % ticks={major also at={1}}
       },
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=xpose/speedup.csv, separator=\space];
   \end{tikzpicture}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Speedup Achieved by \acrshort{com}-Optimal Matrix Transposition]{Speedup
   Achieved by \acrshort{com}-Optimal Matrix Transposition; the CPU time used by the
   straightforward matrix transposition algorithm divided by that used by the optimal one}
   \label{fig:xpose-speedup}
\end{figure}

>>>>

% > In contrast to the external-memory model, algorithms in the cache-oblivious model
% > cannot explicitly manage the cache (issue block-read and block-write requests).  This
% > loss of freedom is necessary because the block and cache sizes are unknown.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf (erikcom), page 5
Cache misses take the place of \acrshortpl{io} \x{used in the \gls{emm}} because
cache-oblivious algorithms don't \alts{manage the cache, read or write cache lines}
explicitly.  This wouldn't be possible since the algorithms know neither the cache nor
the cache line size~\cite[5]{erikcom}.
% > The ideal cache uses the optimal off-line strategy of replacing the cache line whose
% > next access is furthest in the future
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1
Instead, the \gls{com} uses the optimal replacement strategy of evicting the cache line
that won't be accessed for the longest time in the future (Bélády's Algorithm).  This
\alts{
  is strangely \alts{at odds, out of touch, unrealistic} with real-world caches that don't
  know the future,
  {seems like out-of-touch, theoretical ivory-tower nonsense},
}.
% > The ideal-cache model makes the perhaps-questionable assumptions that there are only
% > two levels in the memory hierarchy, that memory is managed automatically by an optimal
% > cache-replacement strategy, and that the cache is fully associative.  We address these
% > assumptions in Section 6, showing that to a certain extent, these assumptions entail
% > no loss of generality.
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=2
%
% > LRU and FIFO replacement do just as well as optimal replacement up to a constant
% > factor of memory transfers and up a constant factor wastage of the cache.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > The same argument extends to a variety of other replacement strategies.
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46
%
% > We show that algorithms with regular complexity bounds (Equation (7.1)) (including all
% > algorithms heretofore presented) can be ported to less-ideal caches incorporating
% > least-recently-used (LRU) or first-in, first-out (FIFO) replacement policies [24, p.
% > 378].
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=51
%
% > [I]n many cases [the COM] is provably within a constant factor of a more realistic
% > cache's performance.
%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
%
% > [A]s long as the number of memory transfers depends polynomially on the cache size M,
% > then halving M will only affect the running time by a constant factor.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > Intuitively, algorithms that slow down by a constant factor when memory (M) is reduced
% > to half, are called regular.
%      -- Algorithms for Memory Hierarchies, page 196
However, \citeauthor{coa-thesis} proves that for many algorithms\footnote{%
   Those satisfying equation (7.1) in~\cite[46]{coa-thesis}.  If the number of cache
   misses incurred by the algorithm only \say{depends polynomially on the cache size
   \(M\)}, the \alts{equation, condition} is satisfied~\cite[6]{erikcom}.
   % Halving the cache size only increases the number of cache misses by a constant
   % factor?
}
\alts{
   cache misses only increase by a constant factor when switching to a feasible
   replacement strategy,
   it only increases cache misses by a constant factor compared to various feasible
   replacement strategies,
}%
% ~\cite[lemma 12, \pno~10]{coa-paper}.%
% ~\cite[corollary 13, \pno~10]{coa-paper}.%
~\cite[corollary 19, \pno~46]{coa-thesis}.%
% See <http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46>.
\footnote{%
   e.g. LRU, FIFO, and random replacement
}

@@

Kumar works around this by \say{[stopping] the recursion when the problem size becomes
less than a certain block size and then [using] the simple for loop implementation inside
the block}~\cite[199--201]{afmh}.  This seems to \alts{compromise, jeopardize} the
\alts{cache-obliviousness, cache-oblivious nature} of the algorithm.
Perhaps cache-oblivious algorithms should be seen as a starting point for further
optimizations that \alts{explicitly use the available information about the memory
hierarchy of the machine a program runs on, take the properties of the target machine into
account}.

@@

\begin{figure}[hbp]
   \centering
   \input{tex/graphics/xpose-speedup-plot}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Speedup Achieved by \acrshort{com}-Optimal Matrix Transposition]{Speedup
   Achieved by \acrshort{com}-Optimal Matrix Transposition; the CPU time used by the
   straightforward matrix transposition algorithm divided by that used by the optimal one}
   \label{fig:xpose-speedup}
\end{figure}

% \begin{figure}[hbp]
%    \centering
%    \input{tex/graphics/xpose-speedup-complicated-plot}
%    \captionsetup{width=.9\linewidth} % FIXME: hacks!
%    \caption{TODO}
%    \label{fig:xpose-speedup-complicated}
% \end{figure}

### 2017-06-27 tex/methodology.tex
Expand, revise, tweak, and run some measurements

All programs were compiled and run on Linux using the \gls{gcc} with
\mintinline{text}{-O2} (unless noted otherwise) and \mintinline{text}{-march=native}.
% Additionally,
I disabled CPU frequency scaling with \mintinline{text}{cpupower(1)},
% See <https://wiki.archlinux.org/index.php/CPU_frequency_scaling>.
reduced the number of running processes,%
\footnote{%
   I stopped Firefox (which added visible noise) and the X server, for example.
}
% (e.g. leaving Firefox open in the background added lots of noise),
and assigned a high priority to the benchmark process with the \mintinline{text}{chrt(1)}
command.

>>>>

All programs were compiled and run on Linux using the \gls{gcc}\footnote{%
   in version 6.3.1
}
with \mintinline{text}{-O2} or \mintinline{text}{-O3} and
\mintinline{text}{-march=native}.
% Additionally,
I disabled CPU frequency scaling with \mintinline{text}{cpupower(1)},
% See <https://wiki.archlinux.org/index.php/CPU_frequency_scaling>.
reduced the number of running processes,%
\footnote{%
   I stopped Firefox (which added visible noise) and the X server, for example.
}
% (e.g. leaving Firefox open in the background added lots of noise),
and assigned a high priority to the benchmark process with the \mintinline{text}{chrt(1)}
command.
